---
title: "Stat 33A - Lecture Notes 9"
date: Oct 18, 2020
output: pdf_document
---



Apply Function Basics
=====================

Doing the same operation repeatedly is a common pattern in programming.

Vectorization is one way, but not all functions are vectorized.


In R, the "apply functions" are another way to do something repeatedly.

The apply functions call a function on each element of a vector or list.



The `lapply()` Function
---------------------

The first and most important apply function is `lapply()`. The syntax is:
```
lapply(X, FUN, ...)
```

The function `FUN` is called once for each element of `X`, with the element as
the first argument. The `...` is for additional arguments to `FUN`, which are
held constant across all calls.


Unrealistic example:
```{r}

```
In practice, it's clearer and more efficient to use vectorization here.


Let's use the dogs data for some realistic examples:
```{r}

```

`lapply()` always returns the result as a list.

"l" for **list** result.



The `sapply()` Function
---------------------

`sapply()` simplifies the result to a vector, when possible.

"s" for **simplified** result.

Examples:
```{r}

```

The `sapply()` function is useful if you are working interactively.






Apply Function Examples
=======================

The California Counties AQI data set is available on the bCourse (`aqi.zip`).

Let's load one of the files:
```{r}

```

What are the classes of the columns?
```{r}

```

How can we load all of the files?
```{r}

```

The `rbind` function combines two data frames by stacking them together:
```{r}

```

The data frames we want to stack are in a list.

How can we call `rbind` on all of them?


The `do.call` function calls a function using a list as the arguments:
```{r}

```

How can we convert multiple columns to a different class?
```{r}

```


Are there any missing values in the columns?
```{r}
colSums(sapply(aqi_df, is.na))
```

How can we compute summary statistics about the numeric columns?
```{r}
is_numeric = sapply(aqi_df, is.numeric)
sapply(aqi_df[is_numeric], mean)
```











The Split-Apply Strategy
========================

The `split()` function splits a vector or data frame into groups based on some
other vector (usually congruent).

```{r}

```


Split weight of dogs by the group column:
```{r}

```

The `split()` function is especially useful when combined with `lapply()` or
`sapply`().

```{r}

```
This is an R idiom!



The `tapply()` Function
---------------------

The `tapply()` function is equivalent to the `split()` and `sapply()` idiom.

"t" for **table**, because `tapply()` is a generalization of the
frequency-counting function `table()`.


Examples:
```{r}

```

This strategy is important for analyzing tabular data regardless of what
programming language or packages you're using.



Split-Apply and dplyr
=====================

We'll use the dogs data here:
```{r}

```

The split-apply strategy is often used to compute grouped statistics.

For example, we can compute the mean weight of the dogs by group:
```{r}

```

The `aggregate` function does the same thing as `tapply`, but returns a data
frame:
```{r}

```

The dplyr `group_by` and `summarize` functions are another form of split-apply:
```{r}

```



Choosing an Apply Function
==========================


1. Is the function you want to apply vectorized?

   If yes, use vectorization.

   Otherwise, continue to #2.


2. Do you want to apply the function to elements or to groups?

   For elements, continue to #3.

   For groups, use the split-apply pattern. Use `split()`, then
   continue to #3 to choose an apply function.

   Note `tapply()` is equivalent to `split()` and `sapply()`.


3. Will the function return the same data type for each element?

   If yes, continue to #4.

   Otherwise, use `lapply()`.


4. Are you working interactively?

   If yes, use `sapply()`.

   Otherwise, use `vapply()`.


Other Apply Functions
---------------------

See this StackOverflow Post for a summary:

    https://stackoverflow.com/a/7141669


The purrr and dplyr packages provide Tidyverse alternatives to apply functions.



Conditional Expressions
=======================

Sometimes you'll need code to do different things, depending on a condition.

_If-statements_ provide a way to write conditional code.


For example, suppose we want to greet one person differently from the others:
```{r}

```

Indent code inside of the if-statement by 2 or 4 spaces.

Indentation makes your code easier to read.



The condition has to be a scalar:
```{r}

```

You can chain together if-statements:
```{r}

```

If-statements return the value of the last expression in the evaluated block:
```{r}

```

Curly braces `{ }` are optional for single-line expressions:
```{r}

```

But you have to be careful if you don't use them:
```{r}

```

The `else` block is optional:
```{r}

```

When there's no `else` block, the value of the `else` block is `NULL`:
```{r}

```



The Congruent Vectors Strategy
==============================

If-statements don't work well with vectors.

For example, suppose we want to transform a vector `x` so that:

* Negative elements are set to 0.
* Positive elements are squared.

Using an if-statement doesn't work for this:
```{r}

```


Instead, use congruent vectors:

1. An input vector (or vectors) to use in conditions.

2. An output vector to store the results.

Use the input vector to conditionally assign elements to the output vector.


So:
```{r}

```


Another example:
```{r}

```


The `ifelse()` Function
-----------------------

R also has a vectorized `ifelse()` function.

For example:
```{r}

```

The `ifelse()` function is less efficient than a regular if-statement or the
congruent vectors strategy.
