---
title: "Stat 33A - Lecture Notes 10"
date: Oct 25, 2020
output: pdf_document
---


Functions
=========

A function is like a factory: raw materials (inputs) go in, products (outputs)
come out.

```
         +-------+
-- in -->|   f   |-- out -->
         +-------+
```


For a function:

* _Parameters_ are placeholder variables for inputs.
    + _Arguments_ are actual values assigned to the parameters in a call.
* The _return value_ is the output.
* The _body_ is the code inside.



You can write your own function with the `function` command.

The syntax is:
```
function(parameter1, parameter2, ...) {
  # Your code goes here
  
  # The result on the last line is automatically returned
}
```

For example, let's make a function that detects negative numbers:
```{r}

```

Generally, you should:

* Assign the function to a variable.
    + Exception: _anonymous functions_ in calls to apply functions
    + Functions are actions, so use a verb in the name.

* Indent the body of the function by 2 or 4 spaces.


As another example, let's make a function that detects leap years:
```{r}

```

Curly braces are optional if the body is only one line:
```{r}

```

To exit the function and return a result early, use `return()`.

It only makes sense to use `return()` inside an if-statement!

For example, suppose we want to return `NA` if the argument isn't a number:
```{r}

```
It's idiomatic to only use `return()` when strictly necessary.



To return multiple values, use a vector or list.

For example, let's make a function that computes the mean and median for a
vector:
```{r}

```






Default Arguments
=================

Parameters can have default arguments that are used if no argument is provided.

Use `=` (not `<-`) to assign default arguments:
```{r}

```


Function Example
================



Variables: Scope & Lookup
=======================

Local Variables
---------------

A variable's *scope* is the section of code where it exists and is accessible.

The `exists()` function checks whether a variable is in scope:
```{r}

```

When you create a function, you create a new scope.

Variables defined inside of a function are *local* to the function.


Local variables cannot be accessed from outside:
```{r}

```

Local variables are reset each time the function is called:
```{r}

```


The `local()` function also creates a new scope:
```{r}

```
Rarely used in practice.



Lexical Scoping
---------------

A function can use variables defined outside (non-local), but only if those
variables are in scope where the function was **defined**.

This property is called *lexical scoping*.

For example:
```{r}

```

Variables defined directly in the R console are *global* and available to any
function.


Local variables *mask* (hide) non-local variables with the same name:
```{r}

```


ONE EXCEPTION!

We often call variables:
```{r}

```

In this case, the variable must refer to a function, so R ignores local
variables that aren't.

For example:
```{r}

```


Dynamic Lookup
--------------

Variable lookup happens when a function is **called**, not when it's defined.

This is called *dynamic lookup*.

For example:
```{r}

```


Summary
-------

* Function definitions (or `local()`) create a new scope.

* Local variables
    + Are private
    + Get reset for each call
    + Mask non-local variables (exception: function calls)

* *Lexical scoping*: where a function is **defined** determines which non-local
  variables are in scope.

* *Dynamic lookup*: when a function is **called** determines values of
  non-local variables.



Using Functions
===============

Functions are the building blocks for solving problems.

Take a divide-and-conquer approach, breaking large problems into
smaller steps.

Use a short function for each step. Then it's easier to:

* Test that each step works correctly.
* Modify, reuse, or repurpose a step.

Before you write a function, it's useful to go through several steps:

1. Write down what you want to do, in detail. It can also help to
   draw a picture of what needs to happen.

2. Check whether there's already a built-in function. Google it!

3. Write the code to handle a simple case first. For data science
   problems, use a small dataset at this step.
