---
title: "Stat 33A - Lecture Notes 8"
date: Oct 11, 2020
output: pdf_document
---


dplyr Overview
==============

See the slides.


Subsets with dplyr
==================

The `dplyr` package provides functions for taking subsets.


We'll use the dogs data (`dogs.rds` on the bCourse) for examples:
```{r}

```


Use `slice()` to choose rows by position:
```{r}

```


Use `select()` to choose columns by name or position:
```{r}

```
In dplyr, column names are NOT quoted (like ggplot2, unlike base R).



Use `:` to indicate a range of rows or columns:
```{r}

```

In dplyr, you can use `:` with names (unlike base R):
```{r}

```


Use `-` to exclude rows or columns:
```{r}

```

In dplyr, you can use `-` with names (unlike base R):
```{r}

```


Subsets by Condition
--------------------

Use `filter()` to choose rows that satisfy a condition:
```{r}

```

If you pass multiple conditions to `filter()`, they're combined with the AND
operator `&`:
```{r}

```


Unique Rows
-----------

Use `distinct()` to remove duplicate rows:
```{r}

```



Base R versus dplyr
===================

Compared to dplyr, R's built-in functions are more general.

They work with data frames, lists, vectors, and more.

The dplyr functions only work with data frames.


We'll use the dogs data (`dogs.rds` on the bCourse) for examples:
```{r}

```


The base R equivalent of `slice()` is:
```{r}

```

The base R equivalent of `select()` is:
```{r}

```

What if we want all except a specific named column?

The `match()` function returns the position of the first match for its first
argument in its second argument:
```{r}

```

You can use `match()` to exclude a a named column:
```{r}

```


Subset by Condition
-------------------

The base R equivalent of `filter()` is:
```{r}

```


Unique Rows
-----------

What if we want to remove duplicate rows?
```{r}

```

The `duplicated()` function returns `TRUE` for duplicated elements/rows:
```{r}

```

You can use `duplicated()` to remove duplicated rows:
```{r}

```




Transformations with dplyr
==========================

Besides taking subsets, you can also use dplyr to transform data.

We'll use the dogs data (`dogs.rds` on the bCourse) for examples:
```{r}

```


Use `mutate()` to create a new column:
```{r}

```

You can can do this with base R as well:
```{r}

```


Use `arrange()` to sort a data frame:
```{r}

```

Again, this is something you can also do with base R:
```{r}

```


Aggregation
-----------

_Aggregation_ means reducing data down to a smaller or simpler result.

Point statistics are all examples of aggregation:

* Frequency
* Mean
* Median
* Standard deviation

Use the `summarize()` function to compute an aggregate:
```{r}

```


Groups
------

It's often useful to split data into groups, then aggregate each group.

For example, we might want to compute group means or group frequencies.

Use the `group_by()` function to split data into groups:
```{r}

```

You'll almost always use `group_by()` together with `summarize()`:
```{r}

```


You can use `group_by()` with multiple categorical features:
```{r}

```

The `count()` function is a shortcut for computing frequencies:
```{r}

```

You can count in base R with `table()`:
```{r}

```

Later on, we'll learn about how to do the equivalent of `group_by()` and
`summarize()` in base R.




The Pipe Operator
=================

We'll use the dogs data (`dogs.rds` on the bCourse) for examples:
```{r}

```

Computations that involve multiple steps can be hard to read:
```{r}

```

As we add more steps, the problem gets worse.

One solution is to use variables:
```{r}

```

dplyr provides an alternative solution: the pipe operator `%>%`.

The idea is that:
```
g(x) %>% f
```
Is transformed into:
```
f(g(x))
```

So you can write compound expressions left-to-right:
```{r}

```

By default, the left-hand side becomes the first argument of the right-hand
side.


Technical note: the pipe is provided by the magrittr package, which is
installed/loaded when you install/load dplyr.
