---
title: "Stat 33A - Lecture Notes 8"
date: Oct 11, 2020
output: pdf_document
---


dplyr Overview
==============

See the slides.


Subsets with dplyr
==================

The `dplyr` package provides functions for taking subsets.


We'll use the dogs data (`dogs.rds` on the bCourse) for examples:
```{r}
dogs = readRDS("data/dogs/dogs.rds")
```


Use `slice()` to choose rows by position:
```{r}
# install.packages("dplyr")

library(dplyr)

slice(dogs, 1)

slice(dogs, 3)
```


Use `select()` to choose columns by name or position:
```{r}
select(dogs, 1)

select(dogs, c(1, 2, 3))

select(dogs, breed)

select(dogs, breed, weight, height)

select(dogs, c(breed, weight, height))
```
In dplyr, column names are NOT quoted (like ggplot2, unlike base R).



Use `:` to indicate a range of rows or columns:
```{r}
select(dogs, 1:3)

slice(dogs, 3:13)
```

In dplyr, you can use `:` with names (unlike base R):
```{r}
select(dogs, popularity:price)
```


Use `-` to exclude rows or columns:
```{r}
select(dogs, -1)

select(dogs, -2)
```

In dplyr, you can use `-` with names (unlike base R):
```{r}
select(dogs, -datadog)

select(dogs, -datadog, -group)
```


Subsets by Condition
--------------------

Use `filter()` to choose rows that satisfy a condition:
```{r}
filter(dogs, weight > 10)

filter(dogs, popularity_all <= 10)

filter(dogs, popularity_all <= 10 & price < 1000)

filter(dogs, kids == "medium" | kids == "high")
```

If you pass multiple conditions to `filter()`, they're combined with the AND
operator `&`:
```{r}
filter(dogs, popularity_all <= 10, price < 1000)
```


Unique Rows
-----------

Use `distinct()` to remove duplicate rows:
```{r}
df = data.frame(x = c(1, 1, 3), y = c(1, 1, 4))
df

distinct(df)
```



Base R versus dplyr
===================

Compared to dplyr, R's built-in functions are more general.

They work with data frames, lists, vectors, and more.

The dplyr functions only work with data frames.


We'll use the dogs data (`dogs.rds` on the bCourse) for examples:
```{r}
dogs = readRDS("data/dogs/dogs.rds")
```


The base R equivalent of `slice()` is:
```{r}
library(dplyr)

slice(dogs, 3)

# Base R
dogs[3, ]

slice(dogs, 1:5)

# Base R
dogs[1:5, ]
```

The base R equivalent of `select()` is:
```{r}
select(dogs, 1)
class(select(dogs, 1))

# Base R
dogs[1]
class(dogs[1])

# Slightly different, not equivalent:
dogs[, 1]
class(dogs[, 1])


select(dogs, datadog)

# Base R
dogs["datadog"]

# Slightly different, not equivalent:
dogs[, "datadog"]
dogs$datadog


select(dogs, -breed)

# Base R
dogs[-"breed"]

dogs[-1]
```

What if we want all except a specific named column?

The `match()` function returns the position of the first match for its first
argument in its second argument:
```{r}
match(3, c(4, 3, 6, 2, 3))

match("breed", names(dogs))

match(10, c(4, 3, 6, 2, 3))
```

You can use `match()` to exclude a a named column:
```{r}
dogs[-match("breed", names(dogs))]

#dogs[-match("google", names(dogs))]
```


Subset by Condition
-------------------

The base R equivalent of `filter()` is:
```{r}
filter(dogs, weight > 20 & price < 1000)

# Base R
subset(dogs, weight > 20 & price < 1000)

# Slightly different, not equivalent:
dogs[dogs$weight > 20, ]
```


Unique Rows
-----------

What if we want to remove duplicate rows?
```{r}
df = data.frame(x = c(1, 1, 1, 3), y = c(1, 1, 2, 7))
df

distinct(df)
```

The `duplicated()` function returns `TRUE` for duplicated elements/rows:
```{r}
duplicated(c(1, 1, 2, 1, 3))
```

You can use `duplicated()` to remove duplicated rows:
```{r}
duplicated(df)

df[!duplicated(df), ]
```




Transformations with dplyr
==========================

Besides taking subsets, you can also use dplyr to transform data.

We'll use the dogs data (`dogs.rds` on the bCourse) for examples:
```{r}
dogs = readRDS("data/dogs/dogs.rds")
```


Use `mutate()` to create a new column:
```{r}
library(dplyr)

dogs = mutate(dogs, hwratio = height / weight)
```

You can can do this with base R as well:
```{r}
dogs$hwratio = dogs$height / dogs$weight
dogs
```


Use `arrange()` to sort a data frame:
```{r}
arrange(dogs, weight)

arrange(dogs, desc(weight), height)
```

Again, this is something you can also do with base R:
```{r}
ord = order(dogs$weight, dogs$height)
dogs[ord, ]
```


Aggregation
-----------

_Aggregation_ means reducing data down to a smaller or simpler result.

Point statistics are all examples of aggregation:

* Frequency
* Mean
* Median
* Standard deviation

Use the `summarize()` function to compute an aggregate:
```{r}
summarize(dogs, mean(weight, na.rm = TRUE))

summarize(dogs, sd(price, na.rm = TRUE))

# Base R
mean(dogs$weight, na.rm = TRUE)
```


Groups
------

It's often useful to split data into groups, then aggregate each group.

For example, we might want to compute group means or group frequencies.

Use the `group_by()` function to split data into groups:
```{r}
group_by(dogs, group)
```

You'll almost always use `group_by()` together with `summarize()`:
```{r}
by_group = group_by(dogs, group)
summarize(by_group, mean(weight, na.rm = TRUE))

summarize(group_by(dogs, kids), median(price, na.rm = TRUE))
```


You can use `group_by()` with multiple categorical features:
```{r}
groups = group_by(dogs, kids, grooming)
summarize(groups, max(price, na.rm = TRUE))
```

The `count()` function is a shortcut for computing frequencies:
```{r}
count(dogs, group)

summarize(group_by(dogs, group), n())
```

You can count in base R with `table()`:
```{r}
table(dogs$group)
```

Later on, we'll learn about how to do the equivalent of `group_by()` and
`summarize()` in base R.




The Pipe Operator
=================

We'll use the dogs data (`dogs.rds` on the bCourse) for examples:
```{r}
dogs = readRDS("data/dogs/dogs.rds")
```

Computations that involve multiple steps can be hard to read:
```{r}
library(dplyr)

nrow(filter(dogs, weight < 20))

nrow(filter(dogs, weight < median(dogs$weight, na.rm = TRUE)))
```

As we add more steps, the problem gets worse.

One solution is to use variables:
```{r}
dogs_below = filter(dogs, weight < 20)
nrow(dogs_below)
```

dplyr provides an alternative solution: the pipe operator `%>%`.

The idea is that:
```
g(x) %>% f
```
Is transformed into:
```
f(g(x))
```

So you can write compound expressions left-to-right:
```{r}
# nrow(filter(dogs, weight < 20))

filter(dogs, weight < 20) %>% nrow
```

By default, the left-hand side becomes the first argument of the right-hand
side.


Technical note: the pipe is provided by the magrittr package, which is
installed/loaded when you install/load dplyr.
