---
title: "Stat 33A - Lecture Notes 12"
date: Nov 8, 2020
output: pdf_document
---


While-loops
===========

A while-loop runs a block of code repeatedly while some condition is `TRUE`.

The condition is checked before each iteration.

For example, suppose you want to add up numbers from 0 to 50, but stop as soon
as the total is greater than 50:
```{r}
numbers = seq(0, 50)

total = 0
i = 1
while (total < 50) {
  total = total + numbers[i]
  cat(sprintf("i is %i, total is %i\n", i, total))
  i = i + 1
}

total
i
```

Use a while-loop when you don't know how many iterations are needed until after
the loop runs.


While-loops are a generalization of for-loops.

That means you can write a for-loop as a while-loop:
```{r}

x = c(5, 0, -1, 10)
for (i in seq_along(x)) {
  cat(x[i], "\n")
}

i = 1
while (i <= length(x)) {
  cat(x[i], "\n")
  i = i + 1
}
```
But generally you shouldn't do this!



Developing Iterative Code
=========================

See the slides.


When thinking about writing a loop, try (in order):

1. vectorization
2. apply functions
    * Try an apply function if iterations are independent.
3. for/while-loops
    * Try a for-loop if some iterations depend on others.
    * Try a while-loop if the number of iterations is unknown.
4. recursion
    * Convenient for naturally recursive problems (like Fibonacci),
      but often there are faster solutions.


Before you write the loop, try writing the code for just 1 iteration.

Make sure that code works; it's easy to test code for 1 iteration.

When you have 1 iteration working, then try using the code in a loop
(you will have to make some small changes).

If your loop doesn't work, try to figure out which iteration is
causing the problem. One way to do this is to use message() to print
out information.

Then try to write the code for the broken iteration, get that
iteration working, and repeat this whole process.



Messages, Warnings, and Errors
==============================

The `message`, `warning`, and `stop` functions form R's _condition system_.

The condition system provides a way to report and handle unusual conditions.



The `message` function prints message to the R console:
```{r}
message("Hello")
```


The `warning` function prints a warning:
```{r}
if (c(TRUE, FALSE)) 42

warning("Hi")
```

By default, calling `warning` in a function prints the name of the function:
```{r}
f = function(x, y) {
  warning("This is a warning!")
  x + y
}

f(3, 4)
```

You can use the `call.` parameter to control this:
```{r}
f = function(x, y) {
  warning("This is a warning!", call. = FALSE)
  x + y
}

f(3, 4)
```


The `stop` function prints an error message and stops evaluation:
```{r}
stop("Hi")

f = function(x, y) {
  stop("This is an error!")
  x + y
}

f(3, 4)
```
The `stop` function also has a `call.` parameter.



Preventing Bugs
===============

Modularity
----------

Try to break problems into smaller steps.

Write down the input(s) and output(s) for each step.

Create functions for steps that are reusable. Use parameters for input and the
return value for output.

Be wary of extremely short (1 line) or long (> 20 line) functions. Often a sign
of failure to break down the problem.



Code Clarity
------------

Format your code so that it's easy to read:

* Use whitespace:

    + Put spaces after commas and around binary operators.
    + Indent code inside curly braces `{ }` by 2 spaces or 1 tab.
    + Don't mix tabs and spaces.
    + Separate logical steps or "paragraphs" with blank lines.

* Put closing curly braces `}` on their own line.
    + Exception: use `} else {`

* Use a consistent naming style:
   + `lowercase_with_underscores`
   + `camelCase`

* Use descriptive variable names. Short names are okay for frequently-used
  variables where the context makes the meaning apparent.

* Use comments:
    + To create a big picture plan for how to write your code.
    + To explain tricky code.
    + To summarize the purpose of a "paragraph" of code.
    + To document how to use your functions. Also see roxygen2.


Without formatting:
```{r}
f=function(x,xx){
xx=match.arg(xx,c("celsius","fahrenheit"),several.ok=TRUE)
x1=xx=="fahrenheit"
x[x1]=(x[x1]-32)*5/9
x+273.15}
```

With formatting:
```{r}
to_kelvin = function(temp, unit) {
  unit = match.arg(unit, c("celsius", "fahrenheit"), several.ok = TRUE)
  is_f = unit == "fahrenheit"
  
  temp[is_f] = (temp[is_f] - 32) * 5 / 9
  temp + 273.15
}
```

Defensive Programming
---------------------

Test whether inputs satisfy your assumptions.

For instance, if your function should only work on scalars, check the
lengths of the arguments.

Raise an error with `stop()` if the assumptions aren't satisfied.

Useful functions for testing assumptions:

* `is.X()` functions for testing various properties. For instance,
  `is.character()` to test whether an object is a character vector.

* `inherits()` for testing the class of an object.

* `length()`, `dim()` for testing dimensions.



The R Debugger
==============

Debugging code is the process of confirming, step-by-step, that what you belive
the code does is what the code actually does.

The key idea is to check each step of the code.

Two strategies:

* Work forward through the code from the beginning.
* Work backward from the source of an error.

R has built-in functions to help with debugging.

The `browser()` function pauses the running code and starts R's
debugging system.

For example:
```{r}
f = function(n) {
  total = 0

  for (i in 1:n) {
    #browser()
    total = total + i
  }
  
  total
}

f(10)
```

The key debugger commands are:

* `n` to run the next line
* `s` to "step into" a call
* `c` to continue running the code
* `Q` to quit the debugger
* `where` to print call stack
* `help` to print debugger help


Another example:
```{r}
g = function(x, y) (1 + x) * y

f = function(n) {
  total = 0

  for (i in 1:n) {
    #browser()
    total = total + g(i, i)
  }
  
  total
}

f(11)
```
