---
title: "Stat 33A - Lecture Notes 1"
date: August 23, 2020
output: pdf_document
---

The R Interface
===============

You can interact with and control R by typing code into the prompt.


The R prompt is a Read-Eval-Print Loop (__REPL__):

1. R waits for you to type in some code and press `Enter`.
2. R reads the code.
3. R __evaluates__ the code to compute a result.
4. R prints the result in the console.
5. R loops back around to waiting for you to enter an expression.

A single piece of code is called an __expression__.


```{r}
13 - 7
```


## Getting Help

R has built-in help files.

You can use the `?` command to get help with a specific function:
```{r}
?sin
```
The `?` command should be your first stop when you learn a new function.


To get help with arithmetic commands, you must put the symbol in single or
double quotes:
```{r}
?"+"
```

This works for ordinary functions as well:
```{r}
?"sin"
```

You can use the `??` command to search the help files:
```{r}
??"graphic"
```


## Order of Operations

Arithmetic in R follows an __order of operations__, similar to the order of
operations you probably learned in a math class.

The order from first to last is:

1. Parentheses `( )`
2. Exponentiation `^`
3. Multiplication `*` and Division `/`
4. Addition `+` and Subtraction `-`

Try this out in the R prompt to get a feel for it:
```{r}
5 * 6 - 3

5 * (6 - 3)
```

R has many more operations besides the ones listed above.

You can see R's complete order of operations with the command:
```{r}
?Syntax
```
We'll learn about some of these operations later on.








Variables
=========

A __variable__ is a name for a value.

Use `=` or `<-` to assign a value to a variable:
```{r}
x = 12

y <- 16
```
Either is okay, but choose one and be consistent!

The value can be the result of a computation:
```{r}
x = y

x = 5 * 7 + 3

x = x + 1
```

Variable names can contain letters or numbers, but can't begin with a number:
```{r}
hello14 = 5
```

With variables, you can:

* Reuse the result of a computation
* Write general expressions, such as `a*x + b`
* Break code into small steps, so that it's easier to test and understand.

```{r}
6 * 7 + 3 - 18

step1 = 6 * 7
step2 = step1 + 3 - 18
```


## Copy-on-Write

In R, variables are __copy-on-write__.

That is, if we assign `x` to `y`:
```{r}
x = 13
y = x
```

And then change `x`:
```{r}
x = 16
```

Then `y` remains unchanged:
```{r}
y
```

Originally, `x` and `y` referred to the same value in memory.

When we changed `x` (a "write"), R automatically copied the original value so
that `y` remained the same.



Calls
=====

A **function** is a reusable command that computes something.

The idea is similar to the functions you probably learned about in math class.

To compute the sine of 3, you can write:
```{r}
sin(3)

cos(5)

sum(1, 2, 3)
```

Code that uses a function is said to __call__ that function.

The part of the code where the function is used is also called a __call__.


R has many built in functions for doing math, statistics, and other computing
tasks.
```{r}
log(3)

sqrt(4)
```

## Parameters

A function's inputs are called __parameters__.

Parameters are listed in each function's help file:
```{r}
?sum

?log
```

## Arguments

When you call a function, the values you assign as input are called
__arguments__.

Some functions accept exactly one argument:
```{r}
length(4)

sin(4)
```

Some functions do not require any arguments at all:
```{r}
sessionInfo()
```

Some functions accept multiple arguments:
```{r}
log(10, 2)

log(10)
```

Some functions accept any number of arguments:
```{r}
sum(5, 4, 1)

c(1, 2, 3)
```

You can set arguments based on the order of the parameters:
```{r}
log(10, 2)
```

Or by using the names of the parameters:
```{r}
log(base = 2, x = 10)

log(base = 2, 10)
```

A function has parameters; a call has arguments!
