---
title: "Stat 33A - Lecture Notes 11"
date: Nov 1, 2020
output: pdf_document
---


Printing Output
===============

The `cat` function prints a message to the console:
```{r}

```

Strings lose their quotes and escape sequences are converted.

The `cat` function always returns `NULL`:
```{r}

```


The `print` function prints R's representation of an object to the console:
```{r}

```

Strings keep their quotes and escape sequences are not converted.

The `print` function returns the object being printed:
```{r}

```

The `show` function is similar to `print`, but always returns `NULL`.

A later lecture might cover why both `print` and `show` are necessary.


For most output, `print` is the wrong function to use!


Formatting Output
-----------------

_Escape sequences_ allow you to write characters that aren't on your keyboard
in strings.

Escape sequences always begin with a backslash.

See `?Quotes` for a complete list.

For example, `\n` is a newline:
```{r}

```

The `paste` function combines strings:
```{r}

```

The `sprintf` function can substitute values into strings.

Substitutions are marked with `%` followed by a character:
```{r}

```



For-loops
=========

A for-loop runs a block of code once for each element of a vector or list:
```{r}

```
Curly braces `{ }` are only required if you have multiple lines of code.


You have to write code to save the results from the loop.

In other words, create a variable where you can save results.

For example:
```{r}

```

Use for-loops when some (or all) of the iterations depend on results from other
iterations.

If the iterations are not dependent, use one of:

1. Vectorization (because it's faster)
2. Apply functions (because they're idiomatic)

In some cases, you can use vectorization even when the iterations are
dependent. For example:
```{r}

```


As another example, let's use a loop to compute Fibonacci numbers.

Each Fibonacci number is the sum of the previous two. The first few Fibonacci
numbers are:
```
1 1 2 3 5 8 13
```

We can use a loop to compute the first `n` Fibonacci numbers:
```{r}

```

If you want to reuse this code, you could make it a function:
```{r}

```


Break & Next
------------

Use `break` to exit a loop early:
```{r}

```


Use `next` to skip to the next iteration early:
```{r}

```



Loop Indices
============

If you need indices, using `1:n` can cause bugs:
```{r}

```

Use `seq_len(n)` instead:
```{r}

```

Similarly, using `1:length(x)` can cause bugs:
```{r}

```

Use `seq_along(x)` instead:
```{r}

```

More generally, use `seq()` to produce sequences of indices.



Preallocation
=============

_Preallocation_ means allocating memory for results before a computation.

These functions allocate vectors:

* `character()`
* `complex()`
* `numeric()`
* `logical()`

* `vector()`
* `rep()`

Examples:
```{r}

```

Preallocation is especially important for loops:
```{r}

```



Loops Example
=============

Let's simulate a two-dimensional discrete random walk.

In a random walk, at each time step the walker randomly moves 1 unit along the
x-coordinates or the y-coordinates.

The algorithm is:

1. Flip a coin to determine whether walker moves in x or y.
2. Flip a coin to determine whether walker moves `-1` or `+1`.
3. Update the walker's position.
4. Repeat 1-3 for the desired number of steps.

We can do this with a for-loop:
```{r}

```
